A Transdiciplinary approach to developing a 3D spectral numerical coastal-flow model based on passive macroalgal rafting phenenomea to shed light on ambient and stochastic dispersal and connectivity of nearshore and offshore marine communities  around the coast of South Africa

### 2019-10-19 literature_review.Rmd
added more interesting coefficents to particle dispersion modelling section
added sections on dispersal modelling. moved some sections around and slight resturcturing. included template to begin modifying and using in future

---
title: A Transdiciplinary approach to developing a 3D spectral numerical coastal-flow model based on passive macroalgal rafting phenenomea to shed light on ambient and stochastic dispersal and connectivity of nearshore and offshore marine communities  around the coast of South Africa
author: "Ross Coppin"
output:
  word_document:
    fig_caption: yes
    toc: no
  pdf_document:
    toc: no
bibliography: references.bib
---

@@

## Sedimentation

@@

# Ocean and coastal waves

## Introduction

@@

@@

# Currents

@@

# Kelp-drag dynamics

# Buoyancy/drag properties

```{r BF_1, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 3, fig.height = 3, fig.cap='\\label{fig:BF_1}Conceptual model of bouyancy characteristics.'}
BF_1 <- readPNG("figures/BF_1.png")
grid.raster(BF_1)
```

An object will only float if the object density ($\rho$) is less than or equal to the density of the liquid in which the object is submerged, i.e. $\rho_{object} < \rho_{liquid}$. The bouyancy force (*BF*) is the force of the displaced liquid pushing the object in an upwards direction which is equal to the weight of the displaced liquid and is expressed as: $$BF = W_{displaced-liquid}$$ The weight of the displaced liquid can be calculated by multiplying the mass of the liquid by the gravitational constant $g$ and so the weight of the displaced liquid can be calculated by: $$W_{displaced-liquid} = mg$$ The equation is expanded upon by including the calculation for mass, $m = \rho V$, and so now the weight of the displaced liquid can be calculated as: $$mg = \rho_{liquid} V g$$ where $\rho$ represents the density of the liquid and $V$ represents the portion of the object submerged. Since the object is floating, the BF can also be expressed by the weight of the object and as mentioned previously the weight of the displaced liquid can also be expressed by $mg$, which in turn can be written as $W_{displaced-liquid} = W_{object}$. Now through substituting the calculations for weight, volume and the dimensions of the object, the equation can be simplied to: $$x= h\displaystyle\frac{\rho_o}{\rho_l}$$ where $h$ is the overall height of the object, $\rho_o$ represents the density of the object and $\rho_l$ represents the density of the liquid. It is important to note that this equation is only applicable to objects that are floating i.e. $\rho_{object} < \rho_{liquid}$. The portion of a floating object above the water surface can also be calcuated using a similar approach, and instead of calculating the submerged portion ($x$) the calculation will now solve for the fraction above the surface which is represented by $y$ (see figure \ref{fig:BF_1}). The same equation for calculating density ($\rho = \displaystyle\frac{m}{V}$) will also be used in solving for $y$. As mentioned previously, for floating objects the $BF = m_{object}g$ and the BF can also be defined as the weight of the displaced liquid ($BF = W_{displaced-liquid}$) and therefore the weight of the displaced liquid is equal to the weight of the object ($W_{displaced-liquid} = W_{object}$). On this basis equation ?? can be written as $$m_{liquid}g = m_{object}g$$
Substituting the equation for density into equation ?? the equation is now: $$\rho_{o}V_{o}g = \rho_{l}V_{l}g$$ Important to note is that $V_{o}$ represents the volume of displaced liquid i.e not the entire object volume (see figure \ref{fig:BF_1}). The constant $g$ is removed through cancellation and now the equation can be written as: $$\displaystyle\frac{V_l}{V_o} = \displaystyle\frac{\rho_o}{\rho_l}$$ We also now that the volume of the object submerged is equal to the volume of the displaced liquid and the fraction $\displaystyle\frac{V_l}{V_o}$ represents the portion of the object submerged. Therefore to calculate the portion above the surface this must be subtracted by 1, therefore: $$1-\displaystyle\frac{V_l}{V_o} = y$$
and since $\displaystyle\frac{V_l}{V_o} = \displaystyle\frac{\rho_o}{\rho_l}$ the equation can now be solved via density of the liquid and the object: $$y = 1 - \displaystyle\frac{\rho_o}{\rho_l}$$ Solving the equation algebraically it can now be written as:$$ y = \displaystyle\frac{\rho_l - \rho_o}{\rho_l}$$

Once the fractions of submerged and non-submerged portions of the object have been calculated the next calculation relevant to this study would be to calculate the weight needed to cause the object to sink or as least be completely submerged. This will relate to epiphyte loading on kelp rafts in order to determine raft longevity. If a weight is placed on the object in the conceptual model that has enough weight to fully submerge (but not sink) the object then an addition bouyancy force is needed, which will be represented by $BF_2$ which is equal to the additional mass that has been added which can be expressed as $mg$. Based on the previous calculations and the laws of physics we know that: $$W_{object} = mg = BF_2 = W_{add-liq-displaced}$$

The weight of the additional liquid displaced can now be calculated by substituting the density and volume of the additional displaced liquid along with the gravitational constant $g$ which is expressed as: $$W_{AL} = \rho_{l}V_{AL}g$$
where the subscript $AL$ represents the components of *additional liquid displaced*. The $V$ can also represent the change in volume ($\Delta V$) and so the expression can be expanded to: $$W_{AL} = \rho(\Delta V) g$$} where $\Delta V$ represents the change in volume of the object, i.e $y$.

A study by @lebreton2012 proposed a methododology to track floating debris from source to sink based on discriptions of global waste production and ocean surface currents, using a combination of hydrodynamic and particle dispersion modelling. The particle tracking model applied uses a two stage process, (1) a hydrodynamic model solves the equations of motion to characterise the movement of water within the domain of the model, (2) virtual particles are released into the flow field established in the first step, and particles are allowed to move through hydrodynamic forcing. The authors extracted sea surface currents from the  HYCOM/NCODA ocean circulation modelling system. The HYCOM model is forced by the US Navy's Operational Global Atmospheric Prediction System (NOGAPS). The model also takes into account wind stress, wind speed, heat flux, and precipitation; and was looped five times to accuratley represent ocean circulation patterns for 30 years. The authors used the velocity data aquired from HYCOM and coupled it to the Lagrangian particle tracking model Pol3DD which was used to simulate dispersion.

$$u'= u + \frac{\delta tu_t}{2}$$ $$\upsilon'=\upsilon + \frac{\delta t\upsilon_t}{2}$$
and $u$, $\upsilon$ are orthogonal velocity components, $t$ is time, $\delta t$ is the model time step, $u_\chi$, $u_y$, $\upsilon_y$ are the $u$ and $\upsilon$ spatial velocity gradients, and $u_t$, $\upsilon_t$ are temporal gradients. Horizontal diffusion was modelled as a random walk with seperate longitudinal and lateral coefficients set to simulate random turbulence. The distance increments moved by the particle at each time step is calculated as:
$$\Delta X = R_N \sqrt{6E_1\delta t}$$ $$\Delta y = R_N\sqrt{6E_2\delta t}$$
where $R_N$ is a random number in the uniform range (-1,1) and $E_1,E_2$ are the longitudinal and lateral eddy diffusivities, respectivley. No additional wind stress terms were applied to the motion of particles, as this factor was already expressed in the HYCOM hydrodynamic data. Potential extra stress for emerged parts of debris was considered negligible as all floating debris are assumed to be fully submerged. The Pol3DD model was also modified so provide additional information on individual particles such as origin, age, and trajectory information of individual particles. Since the model considered a spherical earth, particles were allowed to re-enter the simulation if the passed around the 'globe'. In this study only the surface layer was considered as the prinicple layer which determines the fate of floating particles i.e anthorpogenic debris.

>>>>

---
title: A Transdiciplinary approach to developing a 3D spectral numerical coastal-flow model based on passive macroalgal rafting phenenomea to shed light on ambient and stochastic dispersal and connectivity of nearshore and offshore marine communities  around the coast of South Africa
author: "by RM Coppin"
output:
  word_document:
    fig_caption: yes
    toc: yes
    reference_docx: chapter_1_template.docx
  pdf_document:
    toc: no
bibliography: references.bib
---

@@

@@

# Ocean and coastal waves
## Introduction

@@

# Currents
## Ocean currents

## Nearshore currents

@@

# Calculating characteristics of drift
## Buoyancy of drifting objects in the oceans

```{r BF_1, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 3, fig.height = 3, fig.cap='\\label{fig:BF_1}Conceptual model of bouyancy characteristics.'}

BF_1 <- readPNG("figures/chapter_1/bf_cube.png")
grid.raster(BF_1)
```

An object will only float if the object density ($\rho$) is less than or equal to the density of the liquid in which the object is submerged, i.e. $\rho_{object} < \rho_{liquid}$. The bouyancy force (*BF*) is the force of the displaced liquid pushing the object in an upwards direction which is equal to the weight of the displaced liquid and is expressed as: $$BF = W_{displaced-liquid}$$ The weight of the displaced liquid can be calculated by multiplying the mass of the liquid by the gravitational constant $g$ and so the weight of the displaced liquid can be calculated by: $$W_{displaced-liquid} = mg$$ The equation is expanded upon by including the calculation for mass, $m = \rho V$, and so now the weight of the displaced liquid can be calculated as: $$mg = \rho_{liquid} V g$$ where $\rho$ represents the density of the liquid and $V$ represents the portion of the object submerged. Since the object is floating, the BF can also be expressed by the weight of the object and as mentioned previously the weight of the displaced liquid can also be expressed by $mg$, which in turn can be written as $W_{displaced-liquid} = W_{object}$. Now through substituting the calculations for weight, volume and the dimensions of the object, the equation can be simplied to: $$x= h\displaystyle\frac{\rho_o}{\rho_l}$$ where $h$ is the overall height of the object, $\rho_o$ represents the density of the object and $\rho_l$ represents the density of the liquid.

It is important to note that this equation is only applicable to objects that are floating i.e. $\rho_{object} < \rho_{liquid}$. The portion of a floating object above the water surface can also be calcuated using a similar approach, and instead of calculating the submerged portion ($x$) the calculation will now solve for the fraction above the surface which is represented by $y$ (see figure \ref{fig:BF_1}). The same equation for calculating density ($\rho = \displaystyle\frac{m}{V}$) will also be used in solving for $y$. As mentioned previously, for floating objects the $BF = m_{object}g$ and the BF can also be defined as the weight of the displaced liquid ($BF = W_{displaced-liquid}$). Therefore, the weight of the displaced liquid is equal to the weight of the object ($W_{displaced-liquid} = W_{object}$). On this basis equation ?? can be written as $$m_{liquid}g = m_{object}g$$ Substituting the equation for density into equation ?? the equation is now: $$\rho_{o}V_{o}g = \rho_{l}V_{l}g$$ Important to note is that $V_{o}$ represents the volume of displaced liquid i.e not the entire object volume (see figure \ref{fig:BF_1}). The constant $g$ is removed through cancellation and now the equation can be written as: $$\displaystyle\frac{V_l}{V_o} = \displaystyle\frac{\rho_o}{\rho_l}$$ We also now that the volume of the object submerged is equal to the volume of the displaced liquid and the fraction $\displaystyle\frac{V_l}{V_o}$ represents the portion of the object submerged. Therefore to calculate the portion above the surface this must be subtracted by 1, therefore: $$1-\displaystyle\frac{V_l}{V_o} = y$$ and since $\displaystyle\frac{V_l}{V_o} = \displaystyle\frac{\rho_o}{\rho_l}$ the equation can now be solved via density of the liquid and the object: $$y = 1 - \displaystyle\frac{\rho_o}{\rho_l}$$ Solving the equation algebraically it can now be written as:$$ y = \displaystyle\frac{\rho_l - \rho_o}{\rho_l}$$

Once the fractions of submerged and non-submerged portions of the object have been calculated the next calculation relevant to this study would be to calculate the weight needed to cause the object to sink or as least be completely submerged. This will relate to epiphyte loading on kelp rafts in order to determine raft longevity. If a weight is placed on the object in the conceptual model that has enough weight to fully submerge (but not sink) the object then an addition bouyancy force is needed, which will be represented by $BF_2$ which is equal to the additional mass that has been added which can be expressed as $mg$. Based on the previous calculations and the laws of physics we know that: $$W_{object} = mg = BF_2 = W_{add-liq-displaced}$$ The weight of the additional liquid displaced can now be calculated by substituting the density and volume of the additional displaced liquid along with the gravitational constant $g$ which is expressed as: $$W_{AL} = \rho_{l}V_{AL}g$$
where the subscript $AL$ represents the components of *additional liquid displaced*. The $V$ can also represent the change in volume ($\Delta V$) and so the expression can be expanded to: $$W_{AL} = \rho(\Delta V) g$$} where $\Delta V$ represents the change in volume of the object, i.e $y$. To estimate the amount of input into the model the authors used a scaled approach to define the release of particles, and where based on previous research performed by @halpern2008.

## Wind effects on drifting objects in the oceans

$$u'= u + \frac{\delta tu_t}{2}$$ $$\upsilon'=\upsilon + \frac{\delta t\upsilon_t}{2}$$
and $u$, $\upsilon$ are orthogonal velocity components, $t$ is time, $\delta t$ is the model time step, $u_\chi$, $u_y$, $\upsilon_y$ are the $u$ and $\upsilon$ spatial velocity gradients, and $u_t$, $\upsilon_t$ are temporal gradients. Horizontal diffusion was modelled as a random walk with seperate longitudinal and lateral coefficients set to simulate random turbulence. The distance increments moved by the particle at each time step is calculated as:
$$\Delta X = R_N \sqrt{6E_1\delta t}$$ $$\Delta y = R_N\sqrt{6E_2\delta t}$$
where $R_N$ is a random number in the uniform range (-1,1) and $E_1,E_2$ are the longitudinal and lateral eddy diffusivities, respectivley.

# Application of particle dispersion models

A study by @lebreton2012 proposed a methododology to track floating debris from source to sink based on discriptions of global waste production and ocean surface currents, using a combination of hydrodynamic and particle dispersion modelling. The particle tracking model applied uses a two stage process, (1) a hydrodynamic model solves the equations of motion to characterise the movement of water within the domain of the model, (2) virtual particles are released into the flow field established in the first step, and particles are allowed to move through hydrodynamic forcing. The authors extracted sea surface currents from the  HYCOM/NCODA ocean circulation modelling system. The HYCOM model is forced by the US Navy's Operational Global Atmospheric Prediction System (NOGAPS). The model also takes into account wind stress, wind speed, heat flux, and precipitation; and was looped five times to accuratley represent ocean circulation patterns for 30 years. The authors used the velocity data aquired from HYCOM and coupled it to the Lagrangian particle tracking model Pol3DD which was used to simulate dispersion. No additional wind stress terms were applied to the motion of particles, as this factor was already expressed in the HYCOM hydrodynamic data. Potential extra stress for emerged parts of debris was considered negligible as all floating debris are assumed to be fully submerged. The Pol3DD model was also modified so provide additional information on individual particles such as origin, age, and trajectory information of individual particles. Since the model considered a spherical earth, particles were allowed to re-enter the simulation if the passed around the 'globe'. In this study only the surface layer was considered as the prinicple layer which determines the fate of floating particles i.e anthorpogenic debris.

@griffin2017 added to the search for flight MH370 by providing estimates of the trajectory of the debris. This was achieved through a combination of an advanced ocean model calibrated with oceanic drifters, and drifters in the form of aircraft parts.

@@
